Minimax optimal Laplacian smoothing

### 2020-10-16 notes/manuscripts/AISTATS/minimax_optimal_laplacian_smoothing.tex
Final draft for AISTATS submission.

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts, amsthm, amssymb}
\usepackage{graphicx}
\usepackage[colorlinks]{hyperref}
\usepackage[parfill]{parskip}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{fullpage}
\usepackage{mathtools}
\usepackage{tikz}

@@

\DeclarePairedDelimiterX{\norm}[1]{\lVert}{\rVert}{#1}

@@

%%% Misc
\newcommand{\grid}{\mathrm{grid}}
\newcommand{\critr}{R_n}
\newcommand{\dx}{\,dx}
\newcommand{\dy}{\,dy}
\newcommand{\dr}{\,dr}
\newcommand{\dxpr}{\,dx'}
\newcommand{\dypr}{\,dy'}
\newcommand{\wt}[1]{\widetilde{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\ol}[1]{\overline{#1}}
\newcommand{\spec}{\mathrm{spec}}
\newcommand{\LE}{\mathrm{LE}}
\newcommand{\LS}{\mathrm{LS}}
\newcommand{\OS}{\mathrm{OS}}
\newcommand{\PLS}{\mathrm{PLS}}

@@

\begin{document}
\title{Minimax optimal Laplacian smoothing}
\author{Alden Green}
\date{\today}
\maketitle

\textbf{Disclaimer:} This largely follows the same setup as in the draft I have written which combines Laplacian smoothing and Laplacian eigenmaps results. That draft --- document name \textit{graph\_regression2} --- should for the moment be considered obsolete.

\section{Introduction}

\textbf{(1)} Laplacian smoothing is a graph-based approach to nonparametric regression.

\begin{itemize}
	\item In the random design nonparametric regression problem, we observe data $(X_1,Y_1),\ldots,(X_n,Y_n)$, where $X_1,\ldots,X_n$ are independent samples from a distribution $P$ supported on a domain $\Xset \subset \Rd$, and 
	\begin{equation}
	\label{eqn:random_design_regression}
	Y_i = f_0(X_i) + \varepsilon_i
	\end{equation}
	with $\varepsilon_i \sim N(0,1)$ independent Gaussian noise. Our goal is to perform statistical inference on the unknown regression function $f_0: \Xset \to \Reals$, by which we mean either (a) \emph{estimating} $f_0$ by $\wh{f}$, an estimator constructed from the data $(X_1,Y_1),\ldots,(X_n,Y_n)$ or (b) simply \emph{testing} whether $f_0 = 0$, i.e whether there is any signal present. 
	\item In graph-based nonparametric regression, we perform the above inferential tasks by first building a neighborhood graph $G_{n,r}$ which captures the geometry of $P$ and $\mc{X}$ in an appropriate sense, and then constructing an estimate $\wh{f}$ which is smooth with respect to the graph $G_{n,r}$. The neighborhood graph $G_{n,r} = ([n],{\bf W})$ is a weighted, undirected graph on vertices $[n] = \{1,...,n\}$, which we associate with the samples $\{X_1,\ldots,X_n\}$. The $n \times n$ weight matrix ${\bf W} = ({\bf W}_{ij})_{ij}$ encodes proximity between pairs of samples; for a kernel function $K: [0,\infty) \to \Reals$ and connectivity radius $r > 0$, the entries $\mathbf{W}_{ij}$ are given by
	\begin{equation*}
	\label{eqn:neighborhood_graph}
	{\bf W}_{ij} = K\Biggl(\frac{\norm{X_i - X_j}_{\Rd}}{r}\Biggr).
	\end{equation*}
	Then the degree matrix ${\bf D}$ is the $n \times n$ diagonal matrix with entries ${\bf D}_{ii} = \sum_{j = 1}^{n}{\bf W}_{ij}$, and the graph Laplacian can be written as
	\begin{equation}
	\label{eqn:graph_Laplacian}
	\Lap_{n,r} = \bf{D} - \bf{W}
	\end{equation}
	\item The Laplacian smoothing estimator $\wh{f}_{\LS}$ \citep{smola2003} is a penalized least squares estimator, given by
	\begin{equation}
	\label{eqn:laplacian_smoothing}
	\wh{f}_{\LS} := \min_{f \in \Reals^n} \biggl\{\sum_{i = 1}^{n}(Y_i - f_i)^2 + \rho \cdot f^T \Lap_{n,r} f \biggr\}
	\end{equation}
	where $\rho > 0$ acts a tuning parameter on the penalty $f^T \Lap_{{n,r}} f$.
	Assuming~\eqref{eqn:laplacian_smoothing} is a reasonable estimator of $f_0$, the squared empirical norm
	\begin{equation}
	T_{\LS} = \frac{1}{n}\Bigl\|\wt{f}_{\LS}\Bigr\|_2^2 \label{eqn:laplacian_smoothing_test}
	\end{equation}
	is in turn a reasonable statistic to assess whether or not $f_0 = 0$. 
\end{itemize}

\textbf{(2.a)} Laplacian smoothing has many advantages compared to other nonparametric regression methods. For instance:

\begin{itemize}
	\item It is fast, easy, and stable to compute. Letting ${\bf Y} = (Y_1,\ldots,Y_n) \in \Reals^n$, it is not hard to see that the minimizer $\wt{f}_{\LS} = (\rho \Lap_{n,r} + I)^{-1}{\bf Y}$ can be computed by solving a linear equation; in practice, one typically chooses $r$ to be small, with the result being that the matrix $\Lap_{n,r}$ is sparse. There exist known fast solvers of exactly this system.
	\item Is is generalizable to non-standard data---for instance, text or images, or really any data modality on which it is possible to define a kernel.
	\item It is easily adaptable to the semi-supervised learning setting. 
\end{itemize}

\textbf{(2.b)} For this reason a considerable body of work has emerged analyzing the consistency of the graph Laplacian $\Lap_{{n,r}}$ as $n \to \infty$ and $r(n) \to 0$. This is meant in various senses:
\begin{itemize}
	\item Pointwise consistency, meaning $\Lap_{n,r}f \to \Delta_Pf$ for an appropriate limiting operator $\Delta_P$.
	\item Spectral consistency, meaning the eigenvalues $\lambda_k(\Lap_{n,r}) \to \lambda_k(\Delta_P)$.
	\item Consistency of norms, meaning $f^T \Lap_{n,f} f \to \langle f,\Delta_Pf \rangle_{\Leb^2(P)}$. 
\end{itemize}
However, the statistical optimality of Laplacian smoothing is still not well understood.

\textbf{(2.c)} Our main contributions lie in this latter direction, and they can all be summarized as follows: Laplacian smoothing methods are minimax optimal over Sobolev spaces. In more detail, we will show that when $f_0$ belongs to the Sobolev ball $H^1(\Xset;M)$:
\begin{itemize}
	\item \textbf{Minimax optimal estimation.} With high probability, $\frac{1}{n}\norm{\wt{f}_{\LS} - f_0}_{2}^2 \lesssim n^{-2/(2 + d)}$.
	\item \textbf{Minimax optimal testing.}
	A level-$\alpha$ test constructed using $T_{\LS}$ has non-trivial power whenever $\norm{f_0}_{\Leb^2(\Xset)}^2 \gtrsim n^{-4/(4 + d)}$. 
	\item \textbf{Manifold adaptivity.}
	If $\mc{X} \subset \Rd$ is a submanifold of dimension $m < d$, both of the aforementioned rates hold with $d$ replaced by $m$.
\end{itemize}
Finally, we will also establish that the Laplacian smoothing estimator can be altered to take advantage of higher-order smoothness assumptions on $f_0$. This last conclusion will hold only under very limited circumstances.

\paragraph{Organization.}

List out what is to come.

\section{Minimax regression over Sobolev spaces}

\textbf{(3)} Sobolev spaces are classical function spaces, over which linear nonparametric regression methods. One notable such method is smoothing splines, to which we give special attention because of their close connection to Laplacian smoothing.

\textbf{(4)} In fact, an alternative way to motivate Laplacian smoothing nonparametric regression methods---as opposed to the advantages mentioned in \textbf{(2.a)}---comes from viewing them as discrete-time versions of smoothing splines. This is made more rigorous by the consistency results of \textbf{(2.b)}. Viewed in this light, an additional advantage of Laplacian smoothing emerges: it automatically adapts to the geometry of $\Xset$ and $p$.

\textbf{(5)} Smoothing splines are known to have strong minimax optimal properties. Reasoning by analogy, we might hope that Laplacian smoothing would share these properties.

\section{Graph Sobolev classes}

\textbf{(6)} The minimax optimality of smoothing splines rests on two crucial facts: one, the a priori assumption that $|f|_{H^1(\Xset)}$ is bounded; two, an upper bound on the metric entropy of $H^1(\Xset)$, characterized by the decay of its eigenvalues. \textcolor{red}{(TODO)}:(I know this isn't well-stated, and I need to clean it up.)  We now show that qualitatively similar statements hold with respect to

\quad \textbf{(6.1)} the graph Sobolev semi-norm $f_0^T \Lap_{n,r} f_0$ and,

\quad \textbf{(6.2)} the graph eigenvalues $\lambda_k(\Lap_{n,r})$.

\section{Minimax optimality of Laplacian smoothing}

\textbf{(7)} Using \textbf{(6.1)} and \textbf{(6.2)}, we establish upper bounds on the estimation and testing error of Laplacian smoothing, which confirm that Laplacian smoothing is asymptotically minimax rate-optimal over the Sobolev class $H^1(\Xset)$.

\textbf{(8)} While we motivated graph Laplacians by suggesting they might replicate the statistical properties of smoothing splines, we have in fact managed to show something stronger. In particular:

\quad \textbf{(8.1)} When $d \geq 2$, there is no Sobolev embedding of $H^1(\Xset) \subseteq C^s(\Xset)$ for any $s > 0$. Therefore the smoothing spline estimator is ill-posed.

\quad \textbf{(8.2)} Additionally, the metric entropy of $C^1(\Xset)$ is too large to obtain optimal rates for \textcolor{red}{Holder-smoothing splines} when $d \geq 2$. Thus, we see a gap emerge between the optimality properties of the discrete-time (Laplacian smoothing) and continuous-time (smoothing splines) estimators, in favor of the latter.

\quad \textbf{(8.3)} On the other hand, when $d \geq 4$, we no longer get the known optimal rates in the estimation problem. When $d = 4$, our rates are suboptimal by a factor of $\log n$. When $d > 5$ they are very suboptimal. We believe this is related to classical results regarding the entropy of low-smoothness Holder classes, i.e. the results used to justify \textbf{(8.2)}. The interesting point is that the parameter spaces $C^1(\Xset)$ and $H^1(G_{n,r})$ become ``too large'' for different values of $d$.

\quad \textbf{(8.4)} \textcolor{red}{(TODO)}: Translate these results into error in $\Leb^2(P)$ error, if possible.

\subsection{Manifold adaptivity}

\quad \textbf{(9)} We now show that when $\Xset$ is an $m$-dimensional submanifold of $\Reals^d$, Laplacian smoothing approaches automatically ``feel'' the intrinsic dimension $m$ of $\Xset$, which translate into improved upper bounds on estimating and testing error. This confirms the statement in \textbf{(4)} that Laplacian smoothing automatically adapts to the geometry of $\Xset$ and $p$.

\subsection{Higher-order smoothness classes}

\quad \textbf{(10)} Analogously to smoothing splines, Laplacian smoothing can also be adapted to take advantage of additional assumed regularity on $f_0$, i.e. $f_0 \in H^s(\mc{\Xset})$ for $s > 1$. In very limited circumstances, we can show that this adapted estimator achieves the sharper minimax rates over these classes. However, the general story--for all combinations of $s$ and $d$---remains beyond our reach.

\section{Simulations}

\quad \textbf{(11)} Empirically, we demonstrate that the risk for Laplacian smoothing is comparable to that of smoothing splines, in the full dimensional case. Although one cannot carry out a comparison in the case of an unknown manifold --- since the smoothing spline estimator must be changed in a non-obvious manner --- we do show that when the density $p$ is concentrated close to a manifold and $f_0$ is smooth where $p$ is peaked, we have improved performance for $\wh{f}_{\LS}$ compared to $\wh{f}_{\mathrm{SM}}$.

\section{Discussion}

\clearpage

>>>>

\documentclass[twoside]{article}
\usepackage{aistats2021}
\usepackage{amsmath, amsfonts, amsthm, amssymb}
\usepackage{graphicx}
\usepackage[colorlinks]{hyperref}
\usepackage[parfill]{parskip}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{verbatim}

@@

\DeclarePairedDelimiterX{\norm}[1]{\lVert}{\rVert}{#1}
\DeclarePairedDelimiterX{\seminorm}[1]{\lvert}{\rvert}{#1}

% Make a widecheck symbol (thanks, Stack Exchange!)
\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
\DeclareFontShape{U}{mathx}{m}{n}{
	<5> <6> <7> <8> <9> <10>
	<10.95> <12> <14.4> <17.28> <20.74> <24.88>
	mathx10
}{}
\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
\DeclareFontSubstitution{U}{mathx}{m}{n}
\DeclareMathAccent{\widecheck}{0}{mathx}{"71}
% widecheck made

@@

%%% Misc
\newcommand{\grid}{\mathrm{grid}}
\newcommand{\critr}{R_n}
\newcommand{\dx}{\,dx}
\newcommand{\dy}{\,dy}
\newcommand{\dr}{\,dr}
\newcommand{\dxpr}{\,dx'}
\newcommand{\dypr}{\,dy'}
\newcommand{\wt}[1]{\widetilde{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\ol}[1]{\overline{#1}}
\newcommand{\spec}{\mathrm{spec}}
\newcommand{\LE}{\mathrm{LE}}
\newcommand{\LS}{\mathrm{LS}}
\newcommand{\SM}{\mathrm{SM}}
\newcommand{\OS}{\mathrm{OS}}
\newcommand{\PLS}{\mathrm{PLS}}

@@

\begin{document}

\twocolumn[

	\aistatstitle{Minimax Optimal Regression over Sobolev Spaces via Laplacian Regularization on Neighborhood Graphs}

	\aistatsauthor{Alden Green, Sivaraman Balakrishnan, and Ryan J. Tibshirani}

	\aistatsaddress{Carnegie Mellon University}]

\begin{abstract}
	In this paper we study the statistical properties of Laplacian smoothing, a graph-based approach to non-parametric regression. Under standard regularity conditions, we establish upper bounds on the error of the Laplacian smoothing estimator $\wh{f}$, and a goodness-of-fit test based on $\wh{f}$. These upper bounds match the minimax optimal estimation and testing rates of convergence over the first-order Sobolev class $H^1(\Xset)$, where $\Xset \subset \Rd$ and $1 \leq d < 4$; in the estimation problem, they are within a $\log n$ factor of optimal when $d = 4$. Additionally, we prove that Laplacian smoothing is manifold adaptive: if $\Xset$ is an $m$ dimensional manifold embedded in $\Reals^d$ with $m \ll d$, then the error rate of Laplacian smoothing methods depends only on the intrinsic dimension $m$ and not on the ambient dimension $d$. 
\end{abstract}

\section{INTRODUCTION}
In the nonparametric regression problem, we observe samples $(X_1,Y_1),\ldots,(X_n,Y_n)$ according to the signal plus noise model
\begin{equation}
\label{eqn:signal_plus_noise_model}
Y_i = f_{0}(X_i) + \varepsilon_i,
\end{equation}
with $\varepsilon_i \sim N(0,1)$ independent Gaussian noise. Our goal is to perform statistical inference on the unknown regression function $f_0$, by which we mean either \emph{estimating} $f_0$ or more simply \emph{testing} whether $f_0 = 0$, i.e whether there is any signal present.

The Laplacian smoothing estimator $\wh{f}$ \citep{smola2003} is a penalized least squares estimator, defined on a graph. Letting $G = \bigl(\{1,\ldots,n\},W\bigr)$ be a weighted graph over vertices $\{1,\ldots,n\}$ which we associate with the sample points $\{X_1,\ldots,X_n\}$, the estimator $\wh{f}$ is given by
\begin{equation}
\label{eqn:laplacian_smoothing}
\wh{f} :=  \min_{f \in \Reals^n} \biggl\{\sum_{i = 1}^{n}(Y_i - f_i)^2 + \rho \cdot f^T \Lap f \biggr\}.
\end{equation}
Here $\Lap$ is the graph Laplacian matrix (defined in Section~\ref{sec:problem_setup_and_background}), $G$ is typically a geometric graph (such as a $k$NN or neighborhood graph) and the penalty
\begin{equation*}
f^T \Lap f = \frac{1}{2} \sum_{i,j = 1}^{n} W_{ij}\bigl(f_i - f_j\bigr)^2
\end{equation*}
thus encourages $\wh{f}_i \approx \wh{f}_j$ when $X_i \approx X_j$. Assuming~\eqref{eqn:laplacian_smoothing} is a reasonable estimator of $f_0$, the statistic
\begin{equation}
\label{eqn:laplacian_smoothing_test}
\wh{T} := \frac{1}{n}\bigl\|\wh{f}\bigr\|_2^2 
\end{equation}
is in turn a reasonable way to decide if $f_0 = 0$.

Of course there exist many methods for nonparametric regression \citep{gyorfi2006,wasserman2006,tsybakov2008_book}, but Laplacian smoothing has its own set of advantages. For instance:
\begin{itemize}
	\item \emph{Computational ease.} Laplacian smoothing is fast, easy, and stable to compute. The estimate $\wh{f}$ can be computed by solving a symmetric, diagonally dominant linear system. There are known extremely fast and reliable solvers for exactly this problem (see e.g. the seminal papers of \cite{spielman2011,spielman2013,spielman2014}, or the monograph by~\cite{vishnoi2012} and references therein).
	\item \emph{Generality.} Laplacian smoothing is well defined whenever one can associate a graph with some observed responses. This generality lends itself to many different data modalities: for instance text and image classification, to name only two \citep{kondor2002, belkin03a,belkin2006}. 
	\item \emph{Weak supervision.} Although we study Laplacian smoothing in the supervised learning setup~\eqref{eqn:signal_plus_noise_model}, it can be easily adapted to the semi-supervised or unsupervised settings \citep{nadler09,slepcev17,dunlop2020,calder2019b}.
\end{itemize}

For these reasons, a body of work has emerged analyzing the statistical properties of Laplacian smoothing, and graph-based methods more generally. Roughly speaking, these works can be divided into two categories, based on the perspective they adopt. 
\begin{itemize}
	\item In the \emph{fixed design perspective}, the design points $\{X_1,\ldots,X_n\}$ and the graph $G$ are treated as fixed, and inference is carried out on the evaluations $\bigl(f_0(X_1),\ldots,f_0(X_n)\bigr) \in \Reals^n$.\footnote{To be clear, while in this setting $G$ may be a geometric graph, in principle it could be any graph over $\{1,\ldots,n\}$.} In this context, tight upper bounds have been derived on the error of graph-based estimators \citep{wang2016, sadhanala16,sadhanala17,kirichenko2017,kirichenko2018} and tests \citep{sharpnack2013,sharpnack2013b,sharpnack2015}, which certify that such procedures are \emph{optimal} over ``function'' classes $\mc{F} \subseteq \Reals^n$. The downside of this work is that it may be unnatural to assume a priori that the graph $G$ is fixed, or that the evaluations $\bigl(f_0(X_1),\ldots,f_0(X_n)\bigr)$ belong to some discrete ``function'' class.
	\item In the arguably more natural \emph{random design perspective},  one assumes the design points $X_1,\ldots,X_n$ are sampled independently from some distribution $P$ supported on a domain $\Xset \subset \Rd$. Inference is drawn on the regression function $f_0: \Xset \to \Reals$, which is typically assumed to be smooth in some continuum sense, e.g. it possesses a first derivative bounded in $L^{\infty}$ (H\"{o}lder) or $\Leb^2$ (Sobolev) norm. To conduct graph-based inference, the user first builds a neighborhood graph over the random design points---so that $W_{ij}$ is large iff $X_i$ and $X_j$ are close in say Euclidean distance---and then solves e.g.~\eqref{eqn:laplacian_smoothing} or~\eqref{eqn:laplacian_smoothing_test}. In this context, various graph-based procedures have been shown to be \emph{consistent}: as $n \to \infty$, they converge to a \emph{continuum} limit (see \citep{belkin07,trillos2018,vonluxburg2008} among others). However, until recently such statements were not accompanied by error rates, and such error rates as have been proved \citep{lee2016,trillos2020} are not optimal over continuum function spaces, such as H\"{o}lder or Sobolev classes. 
\end{itemize}
Hereafter, we will adopt the random design perspective, and seek to answer the following question:
\begin{quote}{When $f_0$ is assumed to smooth in a continuum sense, does Laplacian smoothing achieve optimal performance for estimation and goodness-of-fit testing?} 
\end{quote}
On the one hand, this is no small question---rather, it is arguably the fundamental question of nonparametric regression, and without an answer one cannot fully compare the statistical properties of Laplacian smoothing to its competitors. On the other hand, it seems difficult to answer: there is a fundamental gap between the \emph{discrete} smoothness imposed by the penalty $f^T \Lap f$ in~\eqref{eqn:laplacian_smoothing} and the \emph{continuum} smoothness assumed on $f_0$, and to obtain sharp upper bounds we will need to bridge this gap in a tight sense.

\section{SUMMARY OF RESULTS}

\paragraph{Advantages of the discrete approach.} In light of this last point, it is worth asking whether there is any \emph{statistical} advantage to solving a discrete problem such as~\eqref{eqn:laplacian_smoothing} (setting aside computational considerations for the moment). After all, we could have instead solved the following variational problem:
\begin{equation}
\label{eqn:thin_plate_spline}
\wt{f} := \argmin_{f} \biggl\{\sum_{i = 1}^{n} \bigl(Y_i - f(X_i)\bigr)^2 + \rho \cdot \int_{\Xset} \|\nabla f(x)\|^2 \,dx \biggr\},
\end{equation}
with the minimum computed over all functions $f$ for which the criterion is finite--or used the statistic
\begin{equation}
\label{eqn:thin_plate_spline_test}
\wt{T} := \|\wt{f}\|_n^2.
\end{equation}
The penalty in~\eqref{eqn:thin_plate_spline} leverages the assumption that $f_0$ possesses a bounded derivative in a seemingly natural manner. Indeed, for this reason the estimator $\wt{f}$ (and to a lesser extent, the statistic $\wt{T}$) are well known--- in the univariate case, $\wt{f}$ is simply a \emph{smoothing spline}, and when $d > 1$ it is an example of a \emph{thin-plate spline}---and their statistical properties are well-understood \citep{vandergeer2000, liu2019}. Moreover, there exist strong connections between smoothing/thin-plate splines and Laplacian smoothing: as we show later on in~\eqref{eqn:seminorm_consistency}, in a certain sense the Laplacian smoothing problem~\eqref{eqn:laplacian_smoothing} can be seen as a discrete and noisy approximation to~\eqref{eqn:thin_plate_spline}. At first blush, this suggests that Laplacian smoothing should at best inherit the statistical properties of smoothing/thin-plate splines, and at worst may have meaningfully larger error.

However, as we shall see the actual story is quite different: remarkably, Laplacian smoothing enjoys optimality properties in cases where the thin-plate spline problem~\eqref{eqn:thin_plate_spline} is not even well-posed. Tables~\ref{tbl:estimation_rates} and~\ref{tbl:testing_rates} summarize the gap. As we establish in Theorems~\ref{thm:laplacian_smoothing_estimation1}-\ref{thm:laplacian_smoothing_testing_manifold}, when computed over an appropriately formed neighborhood graph, Laplacian smoothing estimators and tests are minimax optimal over first-order \emph{continuum} Sobolev balls. This holds true either when $\Xset \subset \Rd$ is a full-dimensional domain and $d = 1,2$ or $3$, or when $\Xset$ is a submanifold of $\Rd$ of intrinsic dimension $m = 1,2,$ or $3$. Additionally, the estimator $\wh{f}$ is nearly minimax optimal (to within a $(\log n)^{1/3}$ factor) when $d$ (or $m$) = $4$. By contrast, smoothing splines are optimal only when $d$ (or $m$) = $1$. In higher dimensions, technically speaking the problem~\eqref{eqn:thin_plate_spline} is not even well-posed, and any practical implementation of the estimator $\wt{f}$ will be inconsistent \citep{green93}. Hence, our results favorably distinguish Laplacian smoothing methods from their natural variational alternatives.
\begin{table}
	\begin{center}
		\begin{tabular}{p{.1\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Dimension & Laplacian smoothing \eqref{eqn:laplacian_smoothing} & Thin-plate splines \eqref{eqn:thin_plate_spline} \\
			\hline
			$d = 1$ & ${\bf n^{-2/3}}$ & \textcolor{red}{${\bf n^{-2/3}}$} \\
			$d = 2,3$ & ${\bf n^{-2/(2 + d)}}$ & \textcolor{red}{$1$} \\
			$d = 4$ & ${\bf n^{-1/3}} (\log n)^{1/3}$ & \textcolor{red}{$1$} \\
			$d \geq 5$  & $(\log n/n)^{4/(3d)}$ &\textcolor{red}{$1$} \\
		\end{tabular}
	\end{center}
	\caption{Summary of estimation rates over first-order Sobolev balls. Rates in black are new results of this paper, and bolded rates are minimax optimal. Although we suppress it for conciseness, in all cases the dependence of the error rate on the radius of the Sobolev ball is also optimal. For thin-plate splines, ``\textcolor{red}{$1$}'' indicates that the estimator is inconsistent. When $\Xset$ is an $m$ dimensional manifold embedded in $\Rd$, all rates hold with $d$ replaced by $m$, without any change in the method in the case of Laplacian smoothing.}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.1\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Dimension & Laplacian smoothing \eqref{eqn:laplacian_smoothing_test} & Thin-plate splines \eqref{eqn:thin_plate_spline_test} \\
			\hline
			$d = 1$ & ${\bf n^{-4/5}}$ & \textcolor{red}{${\bf n^{-4/5}}$} \\
			$d = 2,3$ & ${\bf n^{-4/(4 + d)}}$ & \textcolor{red}{$n^{-1/2}$} \\
			$d \geq 4$\footnotemark & ${ \bf n^{-1/2}}$ & \textcolor{red}{${ \bf n^{-1/2}}$}
		\end{tabular}
	\end{center}
	\caption{Summary of testing rates over fixed radius first-order Sobolev balls; black, red, and bold font used as in Table~\ref{tbl:estimation_rates}. Rates for $d \geq 4$ assume $f_0 \in \Leb^4(\Xset,M)$. When $\Xset$ is an $m$ dimensional manifold embedded in $\Rd$, all rates hold with $d$ replaced by $m$.}
	\label{tbl:testing_rates}
\end{table}
\footnotetext{Assuming $f_0 \in \Leb^4(\Xset,M)$.}
\paragraph{Future extensions.}
To be clear, these results are not the end of the story. For one thing, the estimator $\wh{f}$ is defined only at the design points $\{X_1,\ldots,X_n\}$. As such, we measure loss using the in-sample mean squared error 
\begin{equation*}
\|\wh{f} - f_0\|_n^2 := \frac{1}{n}\sum_{i = 1}^{n} \Bigl(\wh{f}_i - f_0(X_i)\Bigr)^2.
\end{equation*}
In Section~\ref{sec:minimax_optimal_laplacian_smoothing}, we briefly discuss how one could extend $\wh{f}$ to a function over all $\Xset$, in such a way that the out-of-sample mean square error $\|\wh{f} - f_0\|_{\Leb^2(\Xset)}$ remains small, but leave a formal analysis of the problem to future work.

In a different direction, the problem~\eqref{eqn:thin_plate_spline} is only a special, first-order case of thin-plate splines. In general, the degree-$k$ thin plate spline is defined as a minimizer of
\begin{equation*}
\sum_{i = 1}^{n} (Y_i - f(X_i))^2 + \rho \cdot \sum_{|m| = k} \int_{\Xset} \bigl(D^mf(x)\bigr)^2 \,dx
\end{equation*}
where for each multiindex $m$ we write $D^m$ for the partial derivative $D^mf(x):= \partial^kf/\partial x_{1}^{m_1}\ldots \partial x_{d}^{m_d}$. Assuming the $k$th-order partial derivatives $D^mf_0$ are all bounded in $\Leb^2(\Xset)$ norm, the degree-$k$ thin plate spline has error on the order of $n^{-2k/(2k + d)}$ \citep{vandergeer2000}, which is the minimax optimal rate for such functions. Of course, assuming $f_0$ possesses $k$ bounded derivatives for some $2k > d$ is a rather strong assumption which may not be warranted, but at present we do not know whether (adaptations of) Laplacian smoothing on neighborhood graphs can achieve these faster rates.

% \paragraph{Notation.}
% AJG: Need to make sure these are all defined somewhere
% For an integer $p > 0$, we use $\Leb^p(\Xset)$ to refer to the set of functions $f$ for which the $p$th moment $\norm{f}_{\Leb^p(\Xset)}^p := \int_{\Xset} |f(x)|^p \,dx < \infty$, and $C^p(\Xset)$ to refer to those function which are $p$-times continuously differentiable with bounded $p$th derivative. Finally, we use $C,C_1,\ldots$ and $c,c_1,\ldots$ as constants which are independent of $n$ and $f_0$, and for convenience, we will also use asymptotic notation: $a_n \lesssim b_n$ means that $a_n \leq Cb_n$ for some $C > 0$, and $a_n \asymp b_n$ means that $a_n \lesssim b_n$ and $b_n \lesssim a_n$.

\section{BACKGROUND}
\label{sec:problem_setup_and_background}
Before we formally state our main results in Section~\ref{sec:minimax_optimal_laplacian_smoothing}, we define neighborhood graph Laplacians, and review the known minimax rates over first-order Sobolev spaces.

\paragraph{Neighborhood graph Laplacians.}
As mentioned, in the graph-based approach to nonparametric regression we first build a neighborhood graph $G_{n,r} = ([n],W)$ which captures the geometry of $P$ and $\mc{X}$ in an appropriate sense. The $n \times n$ weight matrix $W = (W_{ij})$ encodes proximity between pairs of samples; for a kernel function $K: [0,\infty) \to \Reals$ and connectivity radius $r > 0$, the entries $W_{ij}$ are given by
\begin{equation*}
\label{eqn:neighborhood_graph}
W_{ij} = K\Biggl(\frac{\norm{X_i - X_j}}{r}\Biggr),
\end{equation*}
where $\norm{\cdot}$ is the Euclidean norm in $\Rd$. Then the degree matrix $D$ is the $n \times n$ diagonal matrix with entries $D_{ii} = \sum_{j = 1}^{n}W_{ij}$, and the graph Laplacian can be written as
\begin{equation}
\label{eqn:graph_Laplacian}
\Lap_{n,r} = D - W;
\end{equation}
we add the subscripts $n$ and $r$ to $\Lap$ and $G$ to emphasize that they depend on the random design points $X_1,\ldots,X_n$ and the radius $r$. To avoid any confusion: henceforth when speak of $\wh{f}$ and $\wh{T}$, we will always assume they are fit using the neighborhood graph Laplacian $L = L_{n,r}$.

Note that $L$ is a symmetric, positive semi-definite matrix, and so admits an eigendecomposition $L = \sum_{k = 1}^{n} \lambda_k v_k v_k^T$. We order the eigenvalues $0 = \lambda_1 \leq \lambda_2 \leq \cdots \leq \lambda_n$, and always consider unit-norm eigenvectors $\|v_k\|_2^2 = 1$; we will make reference to these objects later on.

\paragraph{Sobolev spaces.}
Now that we've defined the neighborhood graph Laplacian $\Lap_{n,r}$, we step away from graph-based methods for a moment, to briefly recall some classical results regarding minimax rates over Sobolev classes.

We say that a function $f \in \Leb^2(\Xset)$ belongs to the \emph{first-order Sobolev space} $H^1(\Xset)$ iff for each $j = 1,\ldots,d$, the weak partial derivative $D^jf$ exists and belong to $\Leb^2(\Xset)$. For such functions $f \in H^1(\Xset)$, the Sobolev seminorm $\seminorm{f}_{H^{1}(\Xset)}$ is the average size of the gradient;
\begin{equation*}
\seminorm{f}_{H^1(\Xset)}^2 = \int_{\Xset} \bigl\|\nabla f(x)\bigr\|^2 \,dx
\end{equation*}
with corresponding Sobolev norm $\norm{f}_{H^1(\Xset)} := \norm{f}_{\Leb^2(\Xset)} + |f|_{H^1(\Xset)}$. The Sobolev balls $H^1(\Xset,M)$ for $M > 0$ are then defined as
\begin{equation*}
H^1(\Xset,M) := \Bigl\{f \in H^1(\Xset): \norm{f}_{H^1(\Xset)}^2 \leq M^2\Bigr\}.
\end{equation*}
For further details regarding Sobolev spaces see e.g.~\citep{evans10,leoni2017}.

\paragraph{Minimax rates.}
To carry out a minimax analysis of regression in Sobolev spaces, one must impose some regularity conditions on the design distribution $P$. We shall assume the following.
\begin{enumerate}[label=(P\arabic*)]
	\item
	\label{asmp:domain}
	$P$ is supported on a domain $\Xset \subset \Rd$, an open, connected set with Lipschitz boundary.
	\item
	\label{asmp:bounded_lipschitz_density} 
	$P$ admits a density $p$ bounded away from $0$ and $\infty$, i.e.
	\begin{equation*}
	0 < p_{\min} \leq p(x) \leq p_{\max} < \infty,~~\textrm{for all $x \in \Xset$.}
	\end{equation*}
	Additionally, the restriction of $p$ to $\Xset$ is Lipschitz, with Lipschitz constant $L_p$.
\end{enumerate}

Under~\ref{asmp:domain} and~\ref{asmp:bounded_lipschitz_density}, the minimax estimation rate over Sobolev balls is (see e.g. \citep{tsybakov2008_book}),
\begin{equation}
\label{eqn:sobolev_space_estimation_minimax_rate}
\inf_{\wh{f}} \sup_{f_0 \in H^1(\Xset, M)} \Ebb\Bigl[\norm{\wh{f} - f_0}_{L^2(\Xset)}^2\Bigr] \asymp M^{2d/(2 + d)}n^{-2/(2 + d)}.
\end{equation}

As nonparametric hypothesis testing is (comparatively) less familiar than nonparametric estimation, we briefly summarize the main idea before stating the optimal error rate. In the goodness-of-fit testing problem, we ask for a test function---formally, a Borel measurable function $\phi$ that takes values in $\{0,1\}$--- which can distinguish between the hypotheses
\begin{equation}
\mathbf{H}_0: f_0 = f_0^{\star}, ~~\textrm{versus}~~ \mathbf{H}_a: f_0 \in \mc{F} \setminus \{f_0^{\star}\}.
\end{equation} 
Typically, the null hypothesis $f_0 = f_0^{\star} \in \mc{F}$ reflects the absence of interesting structure, and $\mc{F} \setminus  \{f_0^{\star}\}$ is a set of smooth departures from this null. To fix ideas, as in \cite{ingster2009} we focus on the problem of \emph{signal detection} in Sobolev spaces, where $f_0^{\star} = 0$ and $\mc{F} = H^1(\Xset,M)$ is a first-order Sobolev ball; this is without loss of generality.
